---
title: "T-test_non_tidy_data"
author: "A Bokhobza"
date: "2023-06-16"
output: html_document
---

**Aim:** Perform a T-test on a **non-tidy** dataset.\
This notebook assumes that the data are **NOT** tidy and are saved in a .xlsx file (*i.e:* spreadsheets with one condition per column).\
In addition, it assumes a doubleheader for the conditions to be tested.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Load required libraries

```{r, include=FALSE, warning=FALSE}
library(openxlsx)
library(dplyr)
library(here)
```

## Setup the "[here](https://here.r-lib.org)" package to avoid user-specific path issues

```{r}

here::i_am("Stats_from_non_tidy_data.Rproj")
here()

```

## Load a dataset

Assumes that the data to be analyzed are located within subfolders (data/raw_data) of the cloned directory.\
**To change the file to analyze, modify the Book1.xlsx entry with the desired filename.**

```{r}

df <-read.xlsx(here("data", "raw_data", "some_data.xlsx"), colNames = FALSE, sep.names = "_",  fillMergedCells = TRUE)

```

*Note: if the file is already open elsewhere an error will be thrown, just close the file and try again.*

## Prepare the dataframe before proceeding with the test.

The following lines are:\

-   Creating variables for the headers (**main_headers and sub_headers**).

-   Deleting the column names in the df.

-   Attributing new column names by merging the mian_headers and sub_headers.

-   Converting the data in the df as numeric.

```{r}
main_headers <- as.character(df[1,])
sub_headers <- as.character(df[2,])

df <- df[-c(1,2),]

colnames(df) <- paste(main_headers, sub_headers, sep = "_")

df <- mutate_all(df, as.numeric)

```

## Preliminary tests before performing T.test

To perform a t.test ([paired](http://www.sthda.com/english/wiki/paired-samples-t-test-in-r) or [unpaired](http://www.sthda.com/english/wiki/unpaired-two-samples-t-test-in-r)), one should verify that:

-   the data are following a normal distribution ([Shapiro-Wilk test](http://www.sthda.com/english/wiki/normality-test-in-r))\
-   the variances of the two groups are equal ([Variance F-test](http://www.sthda.com/english/wiki/f-test-compare-two-variances-in-r))

### Shapiro-Wilk test

The function lapply is used to apply the shapiro.test to all the columns of a df (gives a list). The function sapply is used to convert the list in a table.

The Shapiro-Wilk test determines whether the data are following a normal distribution. A significant value (p\<0.05) indicates that the distribution is not normal. In this case the T.test **should not be performed**, but a non-parametric test should be used instead.

```{r}

shapiro_results <- lapply(df, shapiro.test)
shapiro_results <- sapply(shapiro_results, `[`,c("statistic", "p.value"))
write.csv(shapiro_results, here("data", "processed_data", "shapiro_results.csv"))

```

### Variance F-test (specific conditions)

The variance test determines whether the variances of the two samples are equal.

If the test is significant (p value \< 0.05) a **Welch** t-test should be performed.

To modify the conditions to compare modify the two values *df*$* inside the var.test(). *Note: Typing df$ open a auto-completion tab that allows to select the desired condition.\
Do not forget to modify the filename to be saved.

```{r}

f_test_result <- var.test(df$`D3_Nucleus(ss)137`, df$`D3_Nucleus(ds)137`, alternative = "two.sided")

f_test_result_p_value <- data.frame(data.name = gsub("df\\$`|`", "", f_test_result$data.name), p.value = f_test_result$p.value, 
                        stringsAsFactors = FALSE)

write.csv(f_test_result_p_value, here("data", "processed_data", "results_var_test.csv"))


```

## T-test (specific conditions)

Arguments to be modified depending on desired t.test:(Indication from the [help of the t.test function](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/t.test.html))

-   paired = FALSE (default, change to true for paired t.test)

-   var.equal = FALSE (change to true if the results of F

-   test are not significant for both tested variables)

-   conf.level = 0.95 (modify the value as desired)

```{r}

t_test_result <- t.test(df$`D3_Nucleus(ss)137`, df$`D3_Nucleus(ss)500`, paired = FALSE, var.equal = FALSE, conf.level = 0.95)

t_test_result_p_value <- data.frame(data.name = gsub("df\\$`|`", "", t_test_result$data.name), p.value = f_test_result$p.value, 
                        stringsAsFactors = FALSE)

write.csv(t_test_result_p_value, here("data", "processed_data", "t_test_result.csv"))


```


### Tests (F-test and t-test) on all possible combinations
```{r}
# Assuming your dataframe is called 'df'
column_names <- colnames(df)

# Generate combinations of column names
combinations <- combn(column_names, 2, simplify = FALSE)

# Perform var.test for each combination of columns using lapply
results <- lapply(combinations, function(cols) {
  result_name <- paste(cols, collapse = "_vs_")

  # Perform var.test
  result <- var.test(df[[cols[1]]], df[[cols[2]]])

  # Create a data frame with column names and p-value
  result_df <- data.frame(Column1 = cols[1], Column2 = cols[2], p_value = result$p.value)

  return(result_df)
})

# Combine all results into a single data frame
results_df <- do.call(rbind, results)

# Save as .csv file

write.csv(results_df, here("data","processed_data", "var_test_all.csv"))


```

```{r}

results <- lapply(combinations, function(cols) {
  result_name <- paste(cols, collapse = "_vs_")
  
# Perform t.test
  result <- t.test(df[[cols[1]]], df[[cols[2]]])

  # Create a data frame with column names and p-value
  result_df <- data.frame(Column1 = cols[1], Column2 = cols[2], p_value = result$p.value)

  return(result_df)
})

# Combine all results into a single data frame
results_df <- do.call(rbind, results)

# Save as .csv file

write.csv(results_df, here("data","processed_data", "t_test_all.csv"))



```



